---
title: "Additional Predictor with Maximum Effect Size"
author: Tingting Zhan
date: today
format: 
  html:
    page-layout: full
    html-math-method: katex
number-sections: true
toc: true
toc-location: left
toc-depth: 4
toc-title: ''
editor: source
bibliography: maxEff.bib
knitr:
  opts_chunk: 
    collapse: true
    comment: "#" 
vignette: >
  %\VignetteIndexEntry{intro}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

# Introduction

The complete vignette of R package **`maxEff`** (v`r packageVersion('maxEff')`) 

> <https://rpubs.com/tingtingzhan/maxEff>

documents the selection of one from many `numeric` predictors for a regression model, to ensure that the additional predictor has the maximum effect size.

The complete vignette exceeds the file size limit allowed on [CRAN](https://cran.r-project.org/package=maxEff). 


# Prerequisite

## Environment

Package **`maxEff`** (v`r packageVersion('maxEff')`) requires R version 4.5.0 (released 2025-04-11) or higher ([macOS](https://cran.r-project.org/bin/macosx/), [Windows](https://cran.r-project.org/bin/windows/base/), [Linux](https://cran.r-project.org/bin/linux/)).

An Integrated Development Environment (IDE), e.g., [RStudio](https://posit.co/download/rstudio-desktop/) [@RStudio] or [Positron](https://positron.posit.co/download.html), is not required, but highly recommended.


```{r}
#| code-fold: true
#| code-summary: "Environment on author's computer"
#| label: author-env
Sys.info()[c('sysname', 'release', 'machine')]
R.version
```


## Dependency

The dependencies of package **`maxEff`** are detailed in the [complete vignette](https://rpubs.com/tingtingzhan/maxEff). 


Package **`maxEff`** requires the *development versions* of the **`spatstat.*`** family of packages [@spatstat15; @spatstat05]. Installation instructions are provided in the [complete vignette](https://rpubs.com/tingtingzhan/groupedHyperframe) of package **`groupedHyperframe`** [@groupedHyperframe, v`r packageVersion('groupedHyperframe')`], Section 4.1. 


## Installation

Packages **`maxEff`** (v`r packageVersion('maxEff')`) and **`groupedHyperframe`** (v`r packageVersion('groupedHyperframe')`) can be installed using the following command.

```{r}
#| eval: false
utils::install.packages('groupedHyperframe')
utils::install.packages('maxEff')
```

## Getting Started

Examples in the [complete vignette](https://rpubs.com/tingtingzhan/hyper_gam) require that the `search` path has

```{r}
#| warning: false
#| eval: false
library(maxEff)
library(groupedHyperframe)
library(survival)
```


## Acknowledgement

This work is supported by National Institutes of Health, U.S. Department of Health and Human Services grants

-   R01CA222847 ([I. Chervoneva](https://orcid.org/0000-0002-9104-4505), [T. Zhan](https://orcid.org/0000-0001-9971-4844), and [H. Rui](https://orcid.org/0000-0002-8778-261X))
-   R01CA253977 (H. Rui and I. Chervoneva).


# References {.appendix}

::: {#refs}
:::
